A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-23 tex/body.tex
more

\begin{rep}\ourparagraph{Related Work.} Barrett\end{rep}%
\begin{conf}Barrett\end{conf}
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work as of 2007. Since then, datatypes have been added not only to CVC4
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and \begin{rep}to \end{rep}a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related are the automatic structural
induction in both kinds of provers \cite{kersani-peltier-2013,cruanes-201x,wand-weidenbach-201x,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus
%\begin{conf}\cite{kozen-1983}.\end{conf}%
%\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}.\end{rep}%
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is useful to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
be an abbreviation for
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

Given a fixed signature~$\Sig$,
the decision procedure for the ground theory of (co)datatypes %$\thD$
determines the
$\thD$-satisfiability of finite sets $\Ec$ of constraints:\ equalities
and disequalities between ground $\Sig$-terms. The
procedure is formulated as a tableau-like calculus.
% whose derivation rules are applied until saturation or contradiction.

To simplify the presentation, we make a few assumptions about
$\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is reasonable
because noncorecursive codatatypes can be seen as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not very restrictive, because in principle each interpreted type in
$\Nondata$ having finite cardinality~$n$
can be replaced by a %n enumeration
datatype with $n$~nullary constructors
\cite{barrett-et-al-2007}.
As for uninterpreted types, since the constraints in $\Ec$
are ground, they cannot entail an upper
bound on the cardinality of any such type, so it is safe to consider these types infinite.

%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

@@

%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$
%are simplified to $t_i$, and moreover assume that all %additional
%constraints added to $\Ec$ are normalized in this manner.
%We present the calculus in three steps.
%In the first step, we compute the bidirectional closure of $\Ec$;
%in the second step, we make inferences based on acyclicity and uniqueness;
%and in the third step, %(when necessary),
%we branch on constructor types for various terms in $\tEc$.
%Following the conventions from~\cite{},
%the derivation rules of our calculus are given in \emph{guarded assignment form},
%where a rule can be applied to $\Ec$ if it meets all of the specified preconditions for $\Ec$.

@@

Not all normal interpretations are models of codatatypes, because
models must contain all possible infinite terms, not only cyclic ones. However,
acyclic infinite values are uninteresting to us, and by monotonicity of ground
constraints it is trivial to extend any normal interpretation with extra
domain elements to obtain a genuine model if desired. %Hence, we focus on normal interpretations.

@@

%The last precondition is admittedly rather technical.
To ensure disequality constraints are satisfied by an interpretation based on $\ValC$,
it suffices that $\ValC$ is injective modulo $\alpha$-equivalence.
This invariant holds initially, and
the last precondition in the above rule ensures that it is maintained.
%More precisely,
The set $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is an overapproximation of the values
that, when assigned to ${\Varec{x}}$, will cause values in the range of $\ValC$ to become $\alpha$-equivalent.
%This precondition is not strictly necessary in practice, 
%as one could compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
%injectivity afterward, repeating until a suitable value is found.
For infinite codatatypes, it is always possible to find fresh values $v$
because $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is a finite set.
%Injectivity is important to satisfy disequality constraints.
%For infinite codatatypes, it is always possible to find suitable values $v$.
%For example,
%if the longest cycle in $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is of length $n$, a fresh value
%can be constructed with a cycle of length $n + 1$.

@@

%\ourparagraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
constraints for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the \XXXL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\, a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-constraints occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\, C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

%\ourparagraph{Extension to Quantified Formulas.}
While the decision procedure is restricted to ground constraints,
in practice we often want to solve problems that feature universal axioms and existential conjectures,
where checking satisfiability requires reasoning about such ground constraints in the presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete instantiation-based methods~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods extend naturally to %quantified
formulas involving datatypes and codatatypes.

@@

To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is less than $u$ based on a term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.

>>>>

\begin{rep}\ourparagraph{Related Work.} Barrett\end{rep}%
\begin{conf}Barrett\end{conf}
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work as of 2007. Since then, datatypes have been added not only to CVC4
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and \begin{rep}to \end{rep}a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related are the automatic structural
induction in both kinds of provers \cite{kersani-peltier-2013,cruanes-201x,wand-weidenbach-201x,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values. The $\mu$-notation is
inspired by the $\mu$-calculus
\begin{conf}\cite{kozen-1983}.\end{conf}
\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}.\end{rep}

@@

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(%\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}}
  \ldots)} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is useful to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
be an abbreviation for
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

Given a fixed signature~$\Sig$,
the decision procedure for the ground theory of (co)datatypes %$\thD$
determines the
$\thD$-satisfiability of finite sets $\Ec$ of literals:\ equalities
and disequalities between ground $\Sig$-terms. The
procedure is formulated as a tableau-like calculus.
% whose derivation rules are applied until saturation or contradiction.

To simplify the presentation, we make a few assumptions about
$\Sig$.
First, all codatatypes in $\Codata$ are corecursive. This is reasonable
because noncorecursive codatatypes can be seen as nonrecursive
datatypes.
Second, all types $\tau \in \Nondata$ have infinite cardinality.
This is not very restrictive, because in principle each interpreted type in
$\Nondata$ having finite cardinality~$n$
can be replaced by a %n enumeration
datatype with $n$~nullary constructors
\cite{barrett-et-al-2007}.
As for uninterpreted types, since the literals in $\Ec$
are ground, they cannot entail an upper
bound on the cardinality of any such type, so it is safe to consider these types infinite.

%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All literals added to $\Ec$ are also implicitly
%normalized in this manner.

@@

%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$
%are simplified to $t_i$, and moreover assume that all %additional
%literals added to $\Ec$ are normalized in this manner.
%We present the calculus in three steps.
%In the first step, we compute the bidirectional closure of $\Ec$;
%in the second step, we make inferences based on acyclicity and uniqueness;
%and in the third step, %(when necessary),
%we branch on constructor types for various terms in $\tEc$.
%Following the conventions from~\cite{},
%the derivation rules of our calculus are given in \emph{guarded assignment form},
%where a rule can be applied to $\Ec$ if it meets all of the specified preconditions for $\Ec$.

@@

Not all normal interpretations are models of codatatypes, because
models must contain all possible infinite terms, not only cyclic ones. However,
acyclic infinite values are uninteresting to us, and by monotonicity of ground
literals it is trivial to extend any normal interpretation with extra
domain elements to obtain a genuine model if desired. %Hence, we focus on normal interpretations.

@@

%The last precondition is admittedly rather technical.
To ensure disequality literals are satisfied by an interpretation based on $\ValC$,
it suffices that $\ValC$ is injective modulo $\alpha$-equivalence.
This invariant holds initially, and
the last precondition in the above rule ensures that it is maintained.
%More precisely,
The set $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is an overapproximation of the values
that, when assigned to ${\Varec{x}}$, will cause values in the range of $\ValC$ to become $\alpha$-equivalent.
%This precondition is not strictly necessary in practice, 
%as one could compute $\nf{\ValC [ {\Varec{x}} \mapsto \MU {\Varec{x}}.\; t ]}$ optimistically and check
%injectivity afterward, repeating until a suitable value is found.
For infinite codatatypes, it is always possible to find fresh values $v$
because $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is a finite set.
%Injectivity is important to satisfy disequality literals.
%For infinite codatatypes, it is always possible to find suitable values $v$.
%For example,
%if the longest cycle in $\BAD_{\smash{\J}}^{\Varec{x}}(\ValC)$ is of length $n$, a fresh value
%can be constructed with a cycle of length $n + 1$.

@@

%\ourparagraph{Integration into {\relax\XXXL($T\negvthinspace$)}.}
The decision procedure is implemented as a \relax{theory solver} of CVC4, that is, a
specialized solver for determining the satisfiability of conjunctions of
literals for its theory.
Given a theory $T = T_1 \mathrel\cup \cdots \mathrel\cup T_n$ and a set of
input clauses $F$ in conjunctive normal form, the \XXXL($T$) procedure
incrementally builds partial assignments of
truth values to the atoms of $F$ such that no clause in $F$ is falsified. We can regard such a
partial assignment as a set $M$ of true literals.
% where $a \in M$ for all atoms $a$ that
%it assigned the value $\top$ and $\neg\, a \in M$ for all atoms that it assigns
%the value $\bot$.
%at a high level,
By a variant of the Nelson--Oppen method \cite{jovanovic2011sharing,nelson-oppen-1979},
each $T_i$-solver %for $i = 1, \ldots, l$  n?
takes as input the union $M_i$
of (1)~the purified form of $T_i$-literals occurring in $M$, where fresh
variables replace terms containing symbols not belonging to~$T_i$;
(2)~additional (dis)equalities between variables of types not belonging to
$T_i$. Each $T_i$-solver
either reports that a subset $C$ of $M_i$ is $T_i$-unsatisfiable, in which case
$\neg\, C$ is added to $F\!$, adds a clause to $F\!$, or does nothing.
When $M$ is a complete assignment for $F\!$, a theory solver can choose to do
nothing only if $M_i$ is indeed $T_i$-satisfiable.

@@

%\ourparagraph{Extension to Quantified Formulas.}
While the decision procedure is restricted to ground literals,
in practice we often want to solve problems that feature universal axioms and existential conjectures,
where checking satisfiability requires reasoning about such ground literals in the presence of universally quantified formulas.
Many SMT solvers, including CVC4, can reason about quantified formulas using incomplete instantiation-based methods~\cite{MouraBjoerner07,ReynoldsTinelliMoura14}.
These methods extend naturally to %quantified
formulas involving datatypes and codatatypes.

@@

To reduce the number of unique literals considered by the calculus, we
compute a normal form for literals as a preprocessing step. In particular, we
replace $u \teq t$ by $t \teq u$ if $t$ is less than $u$ based on a term
ordering, replace $\smash{\const{C}_{\negvthinspace j}( \vec t \,) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )}$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\smash{\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )}$ by $t_k$, and replace
occurrences of discriminators $\smash{\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t \,) )}$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
